Temporal trends (NAMCS)

### 2017-07-07 ms1/namcs_trends.Rmd
namcs trends per capita

@@

@@

```{r drugs_time}
f = function(design, y, d) {
  subset(design, year==y) %>%
    svytotal(as.formula(paste0('~', d)), design=.) %>%
    data_frame(year=y, abx=d, total=., se=as.numeric(SE(.)))
}

g = function(design) {
  crossing(year=2011:2014, desc=top_abx$abx) %>%
    rowwise() %>% do(f(design, .$year, .$desc)) %>% ungroup()
}

@@

```{r drugs_time_plot}
abx_time %>%
  mutate(ci=1.96*se, ymin=total-ci, ymax=total+ci) %>%
  ggplot(aes(x=year, y=total, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3, scales='free_y') +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

It doesn't look by eye that any trends are particularly strong. I used Poisson regressions to see if anything was changing with time. Only levo:

```{r drugs_time_model}
qpglm = function(design, frmla) svyglm(frmla, design=design, family=quasipoisson(link='log'))

@@

```{r levo_plot}
american_population = data_frame(year=2011:2014, n_ppl=c(312.92, 315.22, 317.47, 319.85) * 1e6)

levo_time = abx_time %>%
  filter(abx=='levo') %>%
  left_join(american_population, by='year') %>%
  mutate(total=total/n_ppl*1000, se=se/n_ppl*1000)

levo_time %>%
  mutate(ci=1.96*se, ymin=total-ci, ymax=total+ci) %>%
  ggplot(aes(x=year, y=total, ymin=ymin, ymax=ymax)) +
  geom_point() + geom_line() + geom_linerange() +
  ylim(5, 21) +
  ylab('Levo claims per 1,000 Americans') +
  theme_minimal()
```

```{r levo_ratio}
t1 = levo_time %>% filter(year==2011) %$% total
v1 = levo_time %>% filter(year==2011) %$% se
t2 = levo_time %>% filter(year==2014) %$% total
v2 = levo_time %>% filter(year==2014) %$% se

@@

show_tot = function(tot, v) sprintf('%g (±%g)', round(tot, 2), round(v, 2))
show_pct = function(tot, v) sprintf('%g%% (±%g%%)', round((tot - 1)*100, digits=1), round(sqrt(v)*100, digits=1))
```

Levo consumption increased from `r show_tot(t1, v1)` prescriptions per 1,000 Americans per year to `r show_tot(t2, v2)`. That is an increase of `r show_pct(tr, vr)`, i.e., doubled.

## Presence of this trend in different age groups

I can confirm that levo is the only drug with a strong trend among adults (rather than everyone):

```{r drugs_time_model_adult_table}
abx_time_model_adult %>%
  select(abx, estimate, p.value) %>%
  mutate(adj.p.value=p.adjust(p.value, method='BH')) %>%
  arrange(p.value) %>%
  kable
```

And, in fact, this trend exists just among the elderly, although it's only marginally significant if you tested for everything.

```{r levo_elderly}
design %>% subset(age_group=='elderly') %>%
  qpglm(levo~year) %>% tidy
```

Levo may be on the decline for children, however:

```{r levo_children}
design %>% subset(age_group=='child') %>%
  qpglm(levo~year) %>% tidy
```

>>>>

census = read_tsv('../../../../db/census/census_totals.tsv')

@@

## Trends in numbers of visits and number of Americans

The number of visits went down during this period:

```{r visits_census}
design %>%
  update(x=1) %>%
  svyby(~x, ~year, design=., svytotal) %>%
  as_data_frame() %>%
  rename(n_visits=x) %>%
  kable
```

But the number of Americans went up (but the number of children went down):

```{r}
census %>% kable
```

@@

### Overall trends

It looks by eye like the total number of antibiotic prescriptions per capita is going down:

```{r overall}
overall_trend = svyby(~n_abx, ~year, design=design, svytotal) %>%
  as_data_frame() %>%
  left_join(census, by='year') %>%
  mutate(abx_per_capita=n_abx/n_total, se_per_capita=se/n_total)

overall_trend %>%
  mutate(ci=1.96*se_per_capita, ymin=abx_per_capita-ci, ymax=abx_per_capita+ci) %>%
  ggplot(aes(x=year, y=abx_per_capita, ymin=ymin, ymax=ymax)) +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

There is no support for a trend per visit:

```{r overall_model1}
svyglm(n_abx ~ year, design=design, family=quasipoisson(link='identity')) %>%
  tidy %>% kable
```

Nor for a trend per capita:

```{r overall_model2}
overall_trend %>%
  mutate(year=year-min(year)) %>%
  lm(abx_per_capita ~ year, weights=1.0/se_per_capita^2, data=.) %>%
  tidy %>% kable
```

### Trends by drug

```{r drugs_time}
# apc = antibiotic per capita
f = function(design, d) {
  svyby(as.formula(paste0('~', d)), ~year, design=design, svytotal) %>%
    as_data_frame() %>%
    rename_(.dots=setNames(d, 'n_abx')) %>%
    mutate(abx=d)
}

g = function(design) {
  lapply(top_abx$abx, function(x) f(design, x)) %>% bind_rows() %>%
    left_join(census, by='year') %>%
    mutate(apc=n_abx/n_total, se=se/n_total)
}

@@

```{r drugs_time_plot}
abx_time %>%
  mutate(ci=1.96*se, ymin=apc-ci, ymax=apc+ci) %>%
  ggplot(aes(x=year, y=apc, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3, scales='free_y') +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

And the same plot again, but with fixed scales:

```{r drugs_time_plot_fixed}
abx_time %>%
  mutate(ci=1.96*se, ymin=apc-ci, ymax=apc+ci) %>%
  ggplot(aes(x=year, y=apc, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3) +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

It doesn't look by eye that any trends are particularly strong.

I used weighted linear regression to look for trends in totals:

```{r drugs_time_totals_model}
abx_total_models = abx_time %>%
  group_by(abx) %>%
  do(tidy(lm(apc ~ year, weight=1.0/se^2, data=.))) %>%
  ungroup()

abx_total_models %>%
  filter(term=='year') %>% select(-term) %>%
  arrange(estimate) %>%
  kable

abx_total_models %>%
  filter(term=='year') %>% select(-term) %>%
  mutate(ci=1.96*std.error, ymin=estimate-ci, ymax=estimate+ci) %>%
  mutate(abx=forcats::fct_reorder(abx, estimate)) %>%
  ggplot(aes(x=abx, y=estimate, ymin=ymin, ymax=ymax)) +
  geom_point() + geom_linerange() +
  coord_flip() +
  theme_minimal()
```

I used Poisson regressions to see if anything was changing *per visit* with time. Only levo:

```{r drugs_time_model}
qpglm = function(design, frmla) {
  svyglm(frmla, design=design, family=quasipoisson(link='identity'))
}

@@

### Levo trend

```{r levo_ratio}
levo_time = abx_time %>%
  filter(abx=='levo') %>%
  arrange(year)

t1 = levo_time %>% filter(year==2011) %$% apc
v1 = levo_time %>% filter(year==2011) %$% se
t2 = levo_time %>% filter(year==2014) %$% apc
v2 = levo_time %>% filter(year==2014) %$% se

@@

show_tot = function(tot, v) sprintf('%g (±%g)', round(tot, 4), round(v, 4))
show_pct = function(tot, v) sprintf('%g%% (±%g%%)', round((tot - 1)*100, digits=1), round(sqrt(v)*100, digits=1))
```

Levo increased from `r show_tot(t1, v1)` to `r show_tot(t2, v2)` per capita, an increase of `r show_pct(tr, vr)`.
